---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# anvltcc

`anvltcc` compiles [anvl](https://github.com/r-xla/anvl) traced array
programs to C kernels through
[tccquickr](https://github.com/sounkou-bioinfo/tccquickr) and the TinyCC
just-in-time compiler from
[Rtinycc](https://github.com/sounkou-bioinfo/Rtinycc). It is a
millisecond-latency CPU compile path for anvl programs with no XLA
dependency: TinyCC compiles the emitted kernel in-process, so the
trace-to-callable round trip suits development loops and environments
where the PJRT/XLA toolchain is unavailable.

The bridge works because anvl's quickr backend already lowers graphs to
declared R — `declare(type(...))` — which is exactly tccquickr's input
dialect. `anvltcc` reuses anvl's own quickr lowering rules wherever they
emit code inside tccquickr's declared subset, and registers bridge-owned
`tinycc` rules for the primitives that have a native tccquickr surface
form: `dot_general` with the standard contraction becomes a `%*%`
contraction nest, and aval-checked broadcasts pass through to tccquickr's
typed scalar-broadcast and recycle accesses.

## From an anvl trace to a running C kernel

```{r trace}
library(anvltcc)

sigmoid_matmul <- function(x, w) {
  1 / (1 + anvl::nv_exp(-anvl::nv_matmul(x, w)))
}
x <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 2)
w <- matrix(c(0.5, -1, 2, 1, 0, 1), nrow = 3)
```

The traced graph lowers to declared R (anvl's rules plus the bridge's
`tinycc` rules), which is also plain R:

```{r lowered}
graph <- anvl::trace_fn(sigmoid_matmul, args = list(
  x = anvl::nv_array(x, backend = "plain"),
  w = anvl::nv_array(w, backend = "plain")
))
lowered <- graph_to_tinycc_r_function(graph)
lowered
all.equal(as.numeric(lowered(x, w)), as.numeric(1 / (1 + exp(-(x %*% w)))))
```

tccquickr plans the loop nests and emits the C kernel:

```{r source}
cat(tinycc_source(sigmoid_matmul, x = x, w = w))
```

TinyCC compiles it in-process, and the callable agrees with R:

```{r jit}
kernel <- tinycc_jit(sigmoid_matmul, x = x, w = w)
all.equal(as.numeric(kernel(x, w)), as.numeric(1 / (1 + exp(-(x %*% w)))))
```

## What is covered

- Elementwise primitives, scalar broadcast, and reductions, through anvl's
  own quickr rules.
- `dot_general` with no batch dimensions and the standard `(2, 1)`
  contraction, through the bridge's `%*%` rule.
- `broadcast_in_dim` whose operand is scalar/length-1 or already the
  target shape, as an aval-checked passthrough.
- Graph constants (closed-over arrays), bound automatically to the
  compiled callable.

A primitive configuration outside this surface falls back to anvl's
quickr rule; when that rule emits loops or mutation, tccquickr refuses
with a structured diagnostic naming the construct, so failures state the
next lowering rule to add rather than miscompiling. Graphs must currently
have one output leaf and no static arguments.

## Benchmarks

See [benchmarks/benchmark.md](benchmarks/benchmark.md) — measured claims
only: compile latency decomposed by pipeline stage, the fused sigmoid
chain against base R under both TinyCC and gcc, and the honest matrix
row where BLAS wins.

## Design

The bridge follows anvl's documented extension pattern: `.onLoad`
registers a `tinycc` interpretation-rule type and attaches rules to
primitives via `primitive[["tinycc"]] <- rule`. The codegen driver walks
the graph once with a two-level rule lookup — `tinycc` first, `quickr`
fallback — so the bridge owns exactly the primitives whose quickr code
would leave tccquickr's declared subset, and anvl's rule coverage does the
rest.
