---
output: github_document
---

<!-- benchmark.md is generated from benchmark.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
```

# anvltcc benchmarks

Measured claims only. All numbers are medians over repeated runs on this
machine at knit time; every table states what each row actually is. Two
non-claims up front: nobody beats BLAS at large plain matrix products
(R's `%*%` calls BLAS, the naive nests here do not compete), and TinyCC
is a non-optimizing compiler, so its code does not generally outrun the
gcc-compiled loops inside base R — the interesting comparisons are the
pipeline stages and the fused whole-kernel paths.

```{r helpers}
library(anvltcc)

median_time <- function(expr, times = 20L) {
  expr <- substitute(expr)
  runs <- vapply(seq_len(times), function(i) {
    t0 <- Sys.time()
    eval.parent(expr)
    as.numeric(difftime(Sys.time(), t0, units = "secs"))
  }, numeric(1))
  stats::median(runs)
}

fmt_s <- function(seconds) {
  if (is.na(seconds)) {
    "n/a"
  } else if (seconds < 1e-3) {
    sprintf("%.0f µs", seconds * 1e6)
  } else if (seconds < 1) {
    sprintf("%.1f ms", seconds * 1e3)
  } else {
    sprintf("%.2f s", seconds)
  }
}

has_quickr <- requireNamespace("quickr", quietly = TRUE)
```

## Compile latency, decomposed

The workload is a traced sigmoid chain. "Full pipeline" is
trace + lowering + tccquickr's typed analysis + backend planning + code
emission (+ compilation for the compiled rows). The tccquickr planning
stage is measured separately via source-only emission, so the difference
between the source row and a compiled row is that row's compile-and-load
step alone.

```{r compile-latency}
sigmoid_chain <- function(a, x, y) 1 / (1 + anvl::nv_exp(-(a * x + y)))

x <- rnorm(1e6)
y <- rnorm(1e6)

lower_once <- function() {
  graph_to_tinycc_r_function(anvl::trace_fn(sigmoid_chain, args = list(
    a = anvl::nv_scalar(2, backend = "plain"),
    x = anvl::nv_array(x, backend = "plain"),
    y = anvl::nv_array(y, backend = "plain")
  )))
}

plan_with <- function(lowered, backend, mode) {
  tccquickr::tccq_plan_backend(
    S7::prop(tccquickr::tccq_analyze(lowered, strict = TRUE), "value"),
    backend,
    tccquickr::tccq_backend_context(mode = mode, target = "c")
  )
}

source_pipeline <- median_time(
  plan_with(lower_once(), tccquickr::tccq_c_backend(), "source"),
  times = 5L
)
tinycc_pipeline <- median_time(
  plan_with(lower_once(), tccquickr::tccq_rtinycc_backend(), "jit"),
  times = 5L
)
gcc_pipeline <- median_time(
  plan_with(lower_once(), tccquickr::tccq_c_backend(), "shared_library"),
  times = 3L
)
quickr_pipeline <- if (has_quickr) {
  median_time({
    f <- anvl::with_backend(
      "quickr",
      anvl::jit(function(a, x, y) 1 / (1 + anvl::nv_exp(-(a * x + y))))
    )
    f(
      anvl::nv_scalar(2, backend = "quickr"),
      anvl::nv_array(x, backend = "quickr"),
      anvl::nv_array(y, backend = "quickr")
    )
  }, times = 3L)
} else {
  NA_real_
}

knitr::kable(data.frame(
  pipeline = c(
    "trace + typed analysis + C emission (no compiler)",
    "same + TinyCC in-process compile",
    "same + gcc via R CMD SHLIB",
    "anvl quickr backend (Fortran toolchain), first call"
  ),
  median = c(
    fmt_s(source_pipeline),
    fmt_s(tinycc_pipeline),
    fmt_s(gcc_pipeline),
    fmt_s(quickr_pipeline)
  )
))
```

The rows decompose: the difference between the first two rows is
TinyCC's in-process compile-and-bind step (tens of milliseconds), and
the difference between the first and third is a gcc `R CMD SHLIB` round
trip. An earlier revision of this table measured ~1.5 s for every
anvltcc row — profiling attributed 91% of that to per-query trait
re-assertion and linear registry scans in tccquickr's call resolution,
which tccquickr has since fixed (18x) by asserting traits once at the
registry constructor boundary and resolving through an op-name index.

## Runtime: fused sigmoid chain over 1e6 doubles

Base R evaluates `1 / (1 + exp(-(a * x + y)))` as five vectorized C
loops with four temporary allocations; the emitted kernel is one fused
loop. The same C source runs under TinyCC (no optimizer) and gcc
(`R CMD SHLIB`), which isolates codegen quality from fusion.

```{r runtime-elementwise}
tinycc_kernel <- S7::prop(
  S7::prop(plan_with(lower_once(), tccquickr::tccq_rtinycc_backend(), "jit"), "value"),
  "products"
)@attrs$callable
gcc_kernel <- S7::prop(
  S7::prop(plan_with(lower_once(), tccquickr::tccq_c_backend(), "shared_library"), "value"),
  "products"
)@attrs$callable

tinycc_run <- median_time(tinycc_kernel(2, x, y))
gcc_run <- median_time(gcc_kernel(2, x, y))
base_r_run <- median_time(1 / (1 + exp(-(2 * x + y))))

knitr::kable(data.frame(
  implementation = c(
    "fused kernel, TinyCC",
    "fused kernel, gcc",
    "base R (5 loops, 4 temporaries)"
  ),
  `median runtime` = c(fmt_s(tinycc_run), fmt_s(gcc_run), fmt_s(base_r_run)),
  check.names = FALSE
))
```

## Runtime: matrix product (the honest row)

A plain `200 x 200` matrix product. R's `%*%` is BLAS; the anvltcc
kernel is a naive triple loop.

```{r runtime-matmul}
m <- matrix(rnorm(4e4), 200)
w <- matrix(rnorm(4e4), 200)

matmul <- function(x, w) anvl::nv_matmul(x, w)
matmul_kernel <- tinycc_jit(matmul, x = m, w = w)

tinycc_matmul <- median_time(matmul_kernel(m, w))
blas_matmul <- median_time(m %*% w)

knitr::kable(data.frame(
  implementation = c("anvltcc kernel (naive nest, TinyCC)", "base R %*% (BLAS)"),
  `median runtime` = c(fmt_s(tinycc_matmul), fmt_s(blas_matmul)),
  check.names = FALSE
))
```

TinyCC-compiled kernels can bind R's in-process BLAS symbols
(`R_ext/BLAS.h`, `dgemm_`), so a future contraction implementation can
call BLAS from inside a fused kernel — one memory pass *and* BLAS
throughput — once tccquickr's registry grows a `blas` implementation
target.
